% Unified Optimizer Model
% Vehicle allocation and charge scheduling (Hexaly-based)
\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage[margin=2.5cm]{geometry}

\title{Unified Optimizer Model\\
\large Vehicle--Route Allocation and Charge Scheduling}
\author{Allocation-v2}
\date{}

\begin{document}
\maketitle

\section{Overview}

The unified optimizer combines \textbf{vehicle--route allocation} and \textbf{charge scheduling} in a single optimization model. It supports three modes:

\begin{itemize}[noitemsep]
  \item \textbf{Allocation only:} Select at most one sequence per vehicle so that as many routes as possible are covered, then maximize total sequence score.
  \item \textbf{Scheduling only:} For pre-allocated routes, decide charging power per vehicle per 30-minute slot to meet route energy (hard) and minimize electricity cost and target-SOC shortfall (soft).
  \item \textbf{Integrated:} Joint optimization with a weighted-sum objective: maximize allocation score minus charging cost (and shortfall penalty).
\end{itemize}

The implementation uses Hexaly (or a greedy fallback when Hexaly is inactive). Slot length is fixed at 0.5\,h, so energy in slot $t$ for vehicle $v$ is $\text{power}_{t,v} \times 0.5$\,kWh.

\section{Sets and Indices}

\begin{itemize}[noitemsep]
  \item $i \in \mathcal{I}$: sequence index (each sequence = one vehicle + ordered list of routes + cost).
  \item $r \in \mathcal{R}$: route ID (all routes in the planning window).
  \item $v \in \mathcal{V}$: vehicle index (for scheduling).
  \item $t \in \mathcal{T}$: time-slot index; $|\mathcal{T}| = T$; slot length $\Delta = 0.5$\,h.
  \item $c \in \mathcal{C}$: charger index; $|\mathcal{C}| = C$ (site chargers).
\end{itemize}

\section{Parameters and Data}

\subsection{Allocation}
\begin{itemize}[noitemsep]
  \item $\mathcal{I}_v \subseteq \mathcal{I}$: sequences that use vehicle $v$.
  \item $\mathcal{I}_r \subseteq \mathcal{I}$: sequences that cover route $r$.
  \item $c_i \in \mathbb{R}$: cost/score of sequence $i$ (higher is better).
  \item $W_{\text{route}} \in \mathbb{R}_+$: weight for route coverage (default $10^2$); one extra route dominates score differences.
\end{itemize}

\subsection{Scheduling}
\begin{itemize}[noitemsep]
  \item $\overline{P}_v$: max charging power (kW) for vehicle $v$ (AC charge rate).
  \item $\overline{E}_v$: max cumulative energy (kWh) that can be delivered to $v$ (battery capacity minus current SOC).
  \item $E^{\text{req}}_{v,k}$: energy (kWh) required by vehicle $v$ by checkpoint $k$ (departure); derived from route cumulative energy and current SOC.
  \item $\tau_{v,k}$: slot index just before departure for checkpoint $k$ of vehicle $v$ (so energy must be met by end of slot $\tau_{v,k}-1$ or equivalent).
  \item $C^{\text{site}}$: site capacity (kW).
  \item $d_t$: site demand forecast (kW) in slot $t$; available for charging: $\max(0, C^{\text{site}} - d_t)$.
  \item $p_t$: electricity price in slot $t$.
  \item $\sigma_t = \gamma_{\text{time}} \cdot (T - t) / T$: synthetic time price (earlier slots cheaper); $\gamma_{\text{time}} = 0.01$ default.
  \item $\text{SOC}^{\text{targ}}_v$: target SOC (kWh) for vehicle $v$ (e.g. 75\% of capacity).
  \item $\lambda$: penalty per kWh shortfall from target SOC (default 0.2).
  \item $A_{t,v} \in \{0,1\}$: 1 if vehicle $v$ is available to charge in slot $t$, 0 otherwise.
\end{itemize}

\subsection{Charger Allocation}
\begin{itemize}[noitemsep]
  \item $\overline{P}_c$: max power (kW) of charger power class $c$ (grouped by max\_power from t\_charger).
  \item $N_c \in \mathbb{Z}_+$: count of physical chargers with power $\overline{P}_c$ (number of chargers in power class $c$).
  \item $\text{DC}_c \in \{0,1\}$: 1 if chargers in power class $c$ are DC, 0 if AC.
  \item $\text{connected}_{v,c} \in \{0,1\}$: 1 if vehicle $v$ is already connected to a charger in power class $c$ (fixed assignment).
  \item $\text{last\_night}_v \in \mathcal{C} \cup \{\text{none}\}$: charger power class used by vehicle $v$ during previous nighttime period (7PM--7AM within 24h); if none, assignment is flexible.
  \item $\text{nighttime}_t \in \{0,1\}$: 1 if slot $t$ falls in nighttime hours (19:00--06:59), 0 otherwise.
\end{itemize}

\textbf{Note:} Chargers are grouped into power classes $(c)$ based on their maximum power and DC/AC type. Each power class has a count $N_c$ representing the number of physical chargers with that power specification.

\subsection{Config (weights)}
\begin{itemize}[noitemsep]
  \item $\alpha$: allocation score weight (default 1).
  \item $\beta$: scheduling cost weight (default 1).
  \item $W_{\text{route}}$: route count weight in allocation term (default $10^2$).
\end{itemize}

\section{Decision Variables}

\subsection{Allocation (when allocation is active)}
\begin{itemize}[noitemsep]
  \item $x_i \in \{0,1\}$: 1 if sequence $i$ is selected.
  \item $y_r \in \{0,1\}$: 1 if route $r$ is covered by at least one selected sequence.
\end{itemize}

\subsection{Scheduling (when scheduling is active)}
\begin{itemize}[noitemsep]
  \item $P_{t,v} \in [0, \overline{P}_v]$: charging power (kW) for vehicle $v$ in slot $t$.
  \item $E_{t,v} \in [0, \overline{E}_v]$: cumulative energy (kWh) delivered to vehicle $v$ by end of slot $t$.
  \item $s_v \in [0, \overline{s}_v]$: shortfall (kWh) from target SOC for vehicle $v$ (soft); $\overline{s}_v = \max(0, \text{SOC}^{\text{targ}}_v - \text{SOC}^{\text{init}}_v)$.
\end{itemize}

\subsection{Charger Allocation (when chargers provided)}
\begin{itemize}[noitemsep]
  \item $z_{v,c} \in \{0,1\}$: 1 if vehicle $v$ is assigned to charger power class $c$ for the planning window (binary assignment).
\end{itemize}

\textbf{Note:} Assignment is to a power class, not individual charger. The specific physical charger within the class is determined by operational constraints and availability.

Energy delivered in slot $t$ for vehicle $v$ is $P_{t,v} \cdot \Delta = 0.5\, P_{t,v}$\,kWh.

\section{Objectives by Mode}

\subsection{Allocation only}
Maximize route coverage (priority) plus sequence scores:
\begin{equation}
  \max \quad W_{\text{route}} \sum_{r \in \mathcal{R}} y_r + \sum_{i \in \mathcal{I}} c_i\, x_i.
\end{equation}

\subsection{Scheduling only}
Minimize charging cost plus shortfall penalty:
\begin{equation}
  \min \quad \sum_{t,v} \big( p_t + \sigma_t \big)\, (0.5\, P_{t,v}) + \lambda \sum_v s_v.
\end{equation}

\subsection{Integrated}
Maximize weighted allocation term minus weighted scheduling term:
\begin{equation}
  \max \quad \alpha\, \Bigl( W_{\text{route}} \sum_r y_r + \sum_i c_i\, x_i \Bigr) - \beta\, \Bigl( \sum_{t,v} (p_t + \sigma_t)\, (0.5\, P_{t,v}) + \lambda \sum_v s_v \Bigr).
\end{equation}

\section{Constraints}

\subsection{Allocation constraints (when allocation is active)}

\begin{enumerate}[label=(A\arabic*)]
  \item \textbf{One sequence per vehicle:} For each vehicle $v$,
    \begin{equation}
      \sum_{i \in \mathcal{I}_v} x_i \le 1.
    \end{equation}
  \item \textbf{Each route at most once:} For each route $r$,
    \begin{equation}
      \sum_{i \in \mathcal{I}_r} x_i \le 1.
    \end{equation}
  \item \textbf{Route covered definition:} For each route $r$ with $\mathcal{I}_r \ne \emptyset$,
    \begin{align}
      y_r &\le \sum_{i \in \mathcal{I}_r} x_i, \\
      \sum_{i \in \mathcal{I}_r} x_i &\le |\mathcal{I}_r|\, y_r,
    \end{align}
    so $y_r = 1$ iff at least one selected sequence covers $r$.
\end{enumerate}

\subsection{Scheduling constraints (when scheduling is active)}

\begin{enumerate}[label=(S\arabic*)]
  \item \textbf{Cumulative energy:} For each vehicle $v$,
    \begin{equation}
      E_{0,v} = 0.5\, P_{0,v}, \qquad E_{t,v} = E_{t-1,v} + 0.5\, P_{t,v} \quad (t \ge 1).
    \end{equation}
  \item \textbf{Route energy (hard):} For each vehicle $v$ and each checkpoint $k$ with slot index $\tau_{v,k}$ and required energy $E^{\text{req}}_{v,k}$,
    \begin{equation}
      E_{\tau_{v,k}-1,v} \ge E^{\text{req}}_{v,k}.
    \end{equation}
  \item \textbf{Site capacity:} For each slot $t$,
    \begin{equation}
      \sum_v P_{t,v} \le \max(0,\, C^{\text{site}} - d_t).
    \end{equation}
  \item \textbf{Max SOC:} For each $v$ and $t$,
    \begin{equation}
      E_{t,v} \le \overline{E}_v.
    \end{equation}
  \item \textbf{Charge rate:} For each $v$ and $t$,
    \begin{equation}
      P_{t,v} \le \overline{P}_v.
    \end{equation}
  \item \textbf{Availability:} If $A_{t,v} = 0$, then
    \begin{equation}
      P_{t,v} = 0.
    \end{equation}
  \item \textbf{Target SOC shortfall (soft):} For each $v$ with $\overline{s}_v > 0$,
    \begin{equation}
      s_v \ge \text{SOC}^{\text{targ}}_v - \text{SOC}^{\text{init}}_v - E_{T-1,v}.
    \end{equation}
    Shortfall is penalized in the objective; no hard constraint.
\end{enumerate}

\subsection{Charger allocation constraints (when chargers provided)}

\begin{enumerate}[label=(C\arabic*)]
  \item \textbf{Fixed charger assignment:} For each vehicle $v$ and charger $c$ where $\text{connected}_{v,c} = 1$,
    \begin{equation}
      z_{v,c} = 1, \qquad z_{v,c'} = 0 \quad \forall c' \ne c.
    \end{equation}
    Vehicle already connected to a charger must keep that assignment.
  
  \item \textbf{One charger per vehicle:} For each vehicle $v$,
    \begin{equation}
      \sum_{c \in \mathcal{C}} z_{v,c} = 1.
    \end{equation}
    Each vehicle is assigned to exactly one charger power class.
  
  \item \textbf{Charger capacity constraint:} For each vehicle $v$ and slot $t$,
    \begin{equation}
      P_{t,v} \le \sum_{c \in \mathcal{C}} \min(\overline{P}_v, \overline{P}_c) \cdot z_{v,c}.
    \end{equation}
    Charging power limited by assigned charger's max power (and vehicle capability).
  
  \item \textbf{Nighttime charger continuity:} For each vehicle $v$ with $\text{last\_night}_v = c^* \in \mathcal{C}$ (not connected but used charger $c^*$ during previous nighttime), if any slot $t$ in planning window has $\text{nighttime}_t = 1$, then
    \begin{equation}
      z_{v,c^*} = 1, \qquad z_{v,c'} = 0 \quad \forall c' \ne c^*.
    \end{equation}
    Prevents charger switching disruption during nighttime hours (7PM--7AM).
  
  \item \textbf{Time-slot charger capacity:} For each time slot $t$ and charger power class $c$, let $N_c$ be the count of physical chargers with power $\overline{P}_c$. Then
    \begin{equation}
      \sum_{v \in \mathcal{V}} z_{v,c} \cdot \mathbb{1}(P_{t,v} > 0) \le N_c.
    \end{equation}
    At each time slot, the number of vehicles actively charging with power class $c$ cannot exceed the physical count of chargers available in that class. This ensures no overbooking of charging infrastructure at any point in time.
\end{enumerate}

\textbf{Note:} Chargers are grouped by power level (power classes). Multiple vehicles can share the same power class at different times (time-multiplexing), but constraint C5 ensures simultaneous usage respects physical charger availability.

\section{Summary of Default Parameters}

\begin{center}
\begin{tabular}{lll}
  \toprule
  Symbol & Meaning & Default \\
  \midrule
  $W_{\text{route}}$ & Route count weight & $10^2$ \\
  $\alpha$ & Allocation score weight & 1.0 \\
  $\beta$ & Scheduling cost weight & 1.0 \\
  $\lambda$ & Target SOC shortfall penalty & 0.2 \\
  $\gamma_{\text{time}}$ & Synthetic time price factor & 0.01 \\
  Target SOC & Percentage of battery & 75\% \\
  $\Delta$ & Slot length (hours) & 0.5 \\
  Nighttime hours & Charger switching restricted & 19:00--06:59 \\
  \bottomrule
\end{tabular}
\end{center}

\section{Time limits (seconds)}

\begin{center}
\begin{tabular}{ll}
  \toprule
  Mode & Time limit \\
  \midrule
  Allocation only & 30 \\
  Scheduling only & 300 \\
  Integrated & 330 \\
  \bottomrule
\end{tabular}
\end{center}

\section{Charger Allocation Features}

The charger allocation implementation adds the following capabilities:

\begin{itemize}[noitemsep]
  \item \textbf{Power class grouping:} Chargers grouped by max\_power and DC/AC type, reducing decision variables from $V \times C$ individual chargers to $V \times P$ power classes (where $P \ll C$ typically).
  \item \textbf{Binary vehicle--charger assignment:} Each vehicle assigned to exactly one charger power class for the planning window (constant across time slots).
  \item \textbf{Fixed connections:} Vehicles already connected to a charger maintain that power class assignment (hard constraint C1).
  \item \textbf{Power-based optimization:} Considers both vehicle charge rate $\overline{P}_v$ and charger power class $\overline{P}_c$; optimizer selects appropriate power class based on vehicle needs.
  \item \textbf{Nighttime switching prevention:} Tracks previous nighttime charger usage; if vehicle used power class $c$ during previous nighttime (7PM--7AM), must continue using $c$ during current nighttime hours (constraint C4).
  \item \textbf{Time-slot capacity enforcement:} At each 30-minute slot, ensures number of vehicles actively charging per power class does not exceed physical charger count $N_c$ (constraint C5 - \emph{critical for physical feasibility}).
  \item \textbf{Time-multiplexing:} Multiple vehicles can share same power class at different times; allows $V > N_c$ vehicles to use $N_c$ chargers across time.
  \item \textbf{Graceful degradation:} Charger allocation constraints only added when site chargers data provided; backward compatible.
\end{itemize}

\textbf{Data sources:}
\begin{itemize}[noitemsep]
  \item Site chargers: \texttt{t\_charger} table (charger\_id, site\_id, max\_power, dc\_flag), grouped by (max\_power, dc\_flag) into power classes.
  \item Vehicle connections: \texttt{VehicleChargeState.charger\_id} (from vehicle state monitor).
  \item Last nighttime charger: Query \texttt{t\_charge\_schedule} for charging during 19:00--06:59 within 24 hours.
  \item Charger power stored: \texttt{t\_charge\_schedule.assigned\_charger\_power\_kw} (populated by optimizer, used in reports).
\end{itemize}

\textbf{Key benefit:} Constraint C5 prevents overbooking—ensures optimizer cannot assign more vehicles to charge simultaneously than there are physical chargers available (e.g., prevents 5 vehicles from trying to use 2 chargers at the same time).

\section{References}

\subsection{Core Implementation}
\begin{itemize}[noitemsep]
  \item Optimizer: \texttt{src/optimizer/unified\_optimizer.py}
  \item Controller: \texttt{src/controllers/unified\_controller.py}
  \item Data models: \texttt{src/models/scheduler.py} (ChargerPowerClass, VehicleChargeState)
  \item Database queries: \texttt{src/database/queries.py} (GET\_SITE\_CHARGERS, INSERT\_CHARGE\_SCHEDULE)
\end{itemize}

\subsection{Documentation}
\begin{itemize}[noitemsep]
  \item Mathematical model: \texttt{docs/unified\_optimizer\_model.tex} (this document)
  \item Optimization overview: \texttt{docs/optimization\_models.md}
  \item Charger power class implementation: \texttt{CHARGER\_POWER\_CLASS\_IMPLEMENTATION.md}
  \item Time-slot capacity constraint: \texttt{TIMESLOT\_CAPACITY\_CONSTRAINT.md}
  \item Database schema updates: \texttt{DATABASE\_SCHEMA\_UPDATE\_SUMMARY.md}
  \item Quick migration guide: \texttt{MIGRATION\_QUICKSTART.md}
\end{itemize}

\subsection{Testing}
\begin{itemize}[noitemsep]
  \item Charger allocation tests: \texttt{tests/test\_charger\_allocation.py}
  \item Time-slot capacity tests: \texttt{tests/test\_charger\_timeslot\_capacity.py}
\end{itemize}

\subsection{Database}
\begin{itemize}[noitemsep]
  \item Charger data: \texttt{t\_charger} (charger\_id, site\_id, max\_power, dc\_flag)
  \item Schedule storage: \texttt{t\_charge\_schedule} (includes assigned\_charger\_power\_kw)
  \item Migration: \texttt{migrations/001\_add\_charger\_power\_to\_schedule.sql}
\end{itemize}

\end{document}
